---
title: "PCA from Rotation to Dimensionality Reduction"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{PCA from Rotation to Dimensionality Reduction}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 5)
library(pcaLab)
```

## Learning goals

This vignette develops PCA from projection geometry rather than from software output. The sequence is: coordinates, rotation, projected variance, maximum-variance direction, orthogonal residual direction, and low-rank reconstruction.

For centered data matrix $X$ and a unit direction $v$, projected scores are

$$t=Xv$$

and their sample variance is

$$\operatorname{Var}(t)=v^T S v, \qquad S=\frac{X^TX}{n-1}.$$

PCA chooses the unit vector that maximizes this quantity. In two dimensions the search can be visualized by rotating the projection direction through angles from 0 to 180 degrees.

```{r}
r <- pca_teach_rotation(angle = 35)
r$equations
v <- pca_teach_variance()
head(v$data)
```

## Variance is a projection-specific quantity

An orthogonal rotation preserves Euclidean distances and total variance, but redistributes variance among coordinate axes. This is why PCA can concentrate information in a small number of coordinates without changing the geometry of the complete rotated representation.

```{r}
fit <- pca_fit(iris[,1:4], scale = TRUE)
fit$variance_explained
fit$cumulative_variance
```

## Dimensionality reduction

Using only the first $k$ components gives a rank-$k$ approximation:

$$\widehat X_k=T_kV_k^T.$$

The omitted component space is the residual space. Reduction is useful only when the retained approximation preserves scientifically relevant structure.

```{r}
t <- pca_teach_reconstruction(fit)
t$data
```

PCA is unsupervised. Group labels displayed later do not define the principal axes.

## Engine availability and the interactive laboratory

`pca_capabilities()` reports which engines are available in the current installation:

```{r}
pca_capabilities()
```

The interactive laboratory `pca_teach_app()` connects rotation, projected variance, eigenvalues, scores, loadings, and reconstruction in a Shiny application. Launch it from an interactive session:

```{r eval=FALSE}
pca_teach_app(launch.browser = FALSE)
```
