---
title: "Preprocessing, Scaling, and Data Audit"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Preprocessing, Scaling, and Data Audit}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 5)
library(pcaLab)
```

## Preprocessing is part of the model

PCA is sensitive to centering, scaling, and transformations. The chosen preprocessing must therefore be recorded and interpreted scientifically rather than selected only to improve visual separation.

```{r}
a <- pca_preprocess(iris[,1:4], center=TRUE, scale=TRUE, na_action="fail")
a$audit
```

## Scaling options

`pcaLab` supports no scaling, unit-variance scaling, Pareto scaling, range scaling, and VAST scaling. Unit-variance scaling changes a covariance PCA into a correlation-like PCA and prevents a high-variance measurement unit from dominating only because of scale.

```{r}
u <- pca_fit(iris[,1:4], scale=FALSE)
z <- pca_fit(iris[,1:4], scale=TRUE)
cbind(unscaled=u$variance_explained[1:3], standardized=z$variance_explained[1:3])
```

## Missingness and constant variables

`pca_doctor()` summarizes missingness, correlation concentration, p/n geometry, a simple nonlinearity warning, covariance conditioning, and conventional-PCA influence diagnostics.

```{r}
pca_doctor(iris[,1:4])
```

A constant variable is not informative for PCA. A scaling factor that is zero is replaced by 1 with a warning so the issue is visible rather than silently discarded.

## Scaling demonstrated step by step

`pca_teach_scaling()` fits covariance and correlation PCA side by side so the effect of unit-variance scaling is visible:

```{r}
ts <- pca_teach_scaling()
cbind(covariance = ts$data$covariance_pca$variance_explained[1:2],
      correlation = ts$data$correlation_pca$variance_explained[1:2])
```
