---
title: "Scores, Loadings, Contributions, and Biplots"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Scores, Loadings, Contributions, and Biplots}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 5)
library(pcaLab)
```

## Scores and loading directions

Scores are observation coordinates in component space. Loadings are the coefficients that define linear component directions.

```{r}
fit <- pca_fit(iris[,1:4], scale=TRUE, ncomp=3)
head(fit$scores)
fit$loadings
```

For correlation PCA, variable-PC correlations satisfy approximately

$$\operatorname{cor}(X_j,PC_k)=\sqrt{\lambda_k}\,v_{jk}.$$

`pcaLab` reports raw loadings separately from correlation loadings.

```{r}
pca_table(fit,"variables",component=1)
```

## Contribution versus quality of representation

Contribution asks how much a variable helps build a given axis. cos² asks how well a variable is represented by that component. These are not interchangeable.

```{r eval=FALSE}
pca_plot(fit,"contributions",component=1)
pca_plot(fit,"cos2",component=1)
pca_plot(fit,"correlation_circle")
```

## Biplot caution

A biplot overlays observation and variable geometry. Its visual interpretation depends on scaling. `pcaLab` defaults to a readable display using correlation loadings when available, but it does not redefine biplot geometry as a hypothesis test.

## Observation diagnostics and supplementary information

`pca_diagnose()` flags observations with large score or orthogonal distances. `pca_supplementary()` projects new observations or extra variables without refitting the model:

```{r}
head(pca_diagnose(fit))
sup <- pca_supplementary(fit,
       observations = iris[1:3, 1:4],
       variables = cbind(ratio = iris$Petal.Length / iris$Petal.Width))
sup$variable_correlations
```

## Teaching objects for scores, loadings, and biplots

The teaching functions return the same quantities with explanatory equations and interpretation steps:

```{r}
ts <- pca_teach_scores(fit)
head(ts$data)
tl <- pca_teach_loadings(fit)
tl$data$cos2[, 1:2]
tb <- pca_teach_biplot(fit, dims = c(1, 2))
dim(tb$data$scores)
dim(tb$data$loadings)
```
