---
title: "Permutation Inference for PCA"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Permutation Inference for PCA}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 5)
library(pcaLab)
```

## Inferential target

PCA itself is descriptive. `pca_test()` adds a permutation reference model by independently permuting each variable across observations, thereby destroying cross-variable association while preserving marginal distributions.

Global statistics include

$$\Psi=\sum_j(\lambda_j-1)^2$$

and

$$\Phi=\sqrt{\frac{\sum_j\lambda_j^2-p}{p(p-1)}}$$

for standardized variables.

```{r}
fit <- pca_fit(iris[,1:4],scale=TRUE)
test <- pca_test(fit,nperm=49,seed=10)
test$global
test$axes
```

## Axis and variable evidence

The same null simulations provide empirical references for eigenvalues, a loading index, and variable-PC correlations. Multiplicity adjustment is applied to variable-level evidence.

```{r}
head(test$variable_index_q)
head(test$variable_correlation_q)
```

Monte Carlo p-values use $(b+1)/(B+1)$ and therefore are never reported as zero.

## Interpretation boundary

These tests ask whether the observed PCA structure exceeds the chosen permutation null. They do not test treatment differences unless a separate design-based inferential model is specified.
