---
title: "Bootstrap Uncertainty and Subspace Stability"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Bootstrap Uncertainty and Subspace Stability}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 5)
library(pcaLab)
```

## Why naive PCA bootstrap is difficult

Bootstrap refits can reverse signs or exchange adjacent components. `pca_boot()` therefore aligns bootstrap loading vectors to the reference solution before summarizing uncertainty.

```{r}
fit <- pca_fit(iris[,1:4],scale=TRUE,ncomp=3)
b <- pca_boot(fit,nboot=49,seed=1)
b$eigenvalue_ci
b$pve_ci
b$sign_stability
```

## Near-degenerate eigenvalues

When eigenvalues are close, individual axes can rotate substantially. `pca_stability()` evaluates nested loading subspaces using principal angles.

```{r}
st <- pca_stability(fit,nboot=49,kmax=3,seed=2)
st$summary
```

A stable two-dimensional subspace with unstable PC1/PC2 directions should be interpreted as a stable plane rather than as two sharply identifiable axes.

## Variable association

Bootstrap intervals and sign stability can be combined with permutation evidence and effect measures.

```{r}
a <- pca_associate(fit,boot=b)
head(a)
```
