---
title: "PCA with Missing Data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{PCA with Missing Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 5)
library(pcaLab)
```

## Missing values change the computational problem

Complete-case deletion can change the target population and waste multivariate information. pcaLab therefore exposes methods that can estimate low-dimensional structure with missing entries.

```{r}
X <- as.matrix(iris[,1:4])
X[sample(length(X),25)] <- NA
n <- pca_fit(X,method="nipals",scale=TRUE,ncomp=2)
e <- pca_fit(X,method="em",scale=TRUE,ncomp=2)
```

## NIPALS

NIPALS alternates score/loading estimates using available entries and deflates the residual matrix. It is useful for moderate missingness but should be checked for convergence and sensitivity.

## EM-style low-rank completion

The internal EM engine initializes missing cells from column means, alternates truncated SVD reconstruction and missing-cell updates, and returns the completed processed matrix together with convergence information.

```{r}
e$extra$converged
e$extra$iterations
```

## Bayesian PCA

If `pcaMethods` is installed, `method="bayesian"` delegates to BPCA. Missing-data uncertainty is model-dependent; imputed values should not be treated as observed measurements.
