---
title: "Casewise Robust PCA"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Casewise Robust PCA}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 5)
library(pcaLab)
```

## Why classical PCA is sensitive to outliers

PCA maximizes variance, so a small number of extreme observations can define the first component. Casewise robust PCA targets contamination where complete rows are atypical.

```{r}
s <- pca_simulate(n=120,p=8,rank=2,noise=.25,casewise=.08,seed=3)
classical <- pca_fit(s$data,scale=TRUE,ncomp=2)
```

When `rospca` is installed:

```{r eval=FALSE}
robust <- pca_fit(s$data,method="robust",scale=TRUE,ncomp=2)
pca_compare(list(classical=classical,robust=robust))
```

The robust engine wraps ROBPCA and retains robust center, score distance, orthogonal distance, and backend diagnostics when available.

## Robustness is not cellwise robustness

If only scattered cells are contaminated, flagging or downweighting entire rows can discard substantial clean information. Use the cellwise engine for that contamination model.
