---
title: "Sparse and Robust Sparse PCA"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Sparse and Robust Sparse PCA}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 5)
library(pcaLab)
```

## Sparse loadings

Ordinary principal components often involve all variables. Sparse PCA imposes a sparsity structure to improve interpretability, but the resulting components need not retain every exact orthogonality/reconstruction property of classical PCA.

```{r eval=FALSE}
sp <- pca_fit(iris[,1:4],method="sparse",scale=TRUE,ncomp=2,backend="sparsepca")
sp$extra$sparsity
```

## Do not reuse classical explained-variance formulas blindly

Different sparse PCA algorithms define scores, loadings, deflation, and explained variance differently. pcaLab retains the backend result and marks sparse fits as non-orthogonal when the ordinary classical identities are not guaranteed.

## Robust sparse PCA

`method="robust_sparse"` adds resistance to outlying observations while favoring sparse loadings. It is distinct from low-rank-plus-sparse RPCA, where “sparse” refers to an error matrix rather than sparse variable loadings.
