---
title: "Clustering on PCA Scores"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Clustering on PCA Scores}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 5)
library(pcaLab)
```

## 1. Why this tutorial exists

PCA reduces high-dimensional data to a small number of principal components. Once the most informative low-dimensional representation is available, clustering on those scores reveals structure that variable-level clustering might obscure.

The central rule is simple:

**Decide the clustering method, the number of clusters, and the degree of stability before interpreting individual groups.**

This tutorial follows that rule from a basic PCA through k-selection, multiple-method comparison, bootstrap stability assessment, cluster boundary computation, custom labelling, and a structured summary.

---

## 2. Learning objectives

After working through this tutorial, the reader should be able to:

1. extract PCA scores and decide how many components to keep;
2. evaluate the number of clusters using silhouette, Calinski-Harabasz, gap statistic, and WSS;
3. compare k-means, PAM, hierarchical clustering, and MCLUST at a fixed k;
4. perform bootstrap resampling to build a consensus matrix and measure membership stability;
5. compute cluster boundaries using Mahalanobis ellipses, convex hulls, or SVM decision surfaces;
6. generate observation, mean, group, and external custom labels for PCA score plots;
7. produce a compact structured report that integrates evaluation, stability, and silhouette;
8. recognize when cluster boundaries are fragile and report that fact clearly.

---

## 3. Package map: clustering functions in pcaLab

| Function | Instructional role | Key arguments |
|---|---|---|
| `pca_cluster_evaluate()` | Evaluate multiple k values across multiple indices | method, k_range, B, dims |
| `pca_cluster()` | Cluster observations using one method at a chosen k | method, k, dims, nstart |
| `pca_cluster_compare()` | Compare multiple methods at the same k | methods, k, dims |
| `pca_cluster_stability()` | Bootstrap resampling for consensus matrix and membership stability | method, k, nboot, dims, seed |
| `pca_cluster_boundary()` | Compute Mahalanobis, convex-hull, or SVM boundaries | method, level, dims |
| `pca_cluster_labels()` | Generate observation, mean, group, or external labels | type, labels, dims |
| `pca_cluster_report()` | Compact multi-part summary integrating all components | method, k_range, dims |

The progression is deliberate. The evaluation panel decides k; the comparison panel decides which method; the stability panel documents how confident the assignment is; the boundary and label panels prepare the final figure; the report brings everything together.

Every function accepts either a `pca_fit` object or a raw numeric matrix. All optional backends (cluster, mclust, dbscan, e1071) are guarded with `requireNamespace()`; missing packages produce informative errors.

---

## 4. The dataset: Fisher's iris measurements

The iris dataset contains 150 observations of 4 morphological measurements from 3 species of iris.

| Species | n | Sepal.Length | Sepal.Width | Petal.Length | Petal.Width |
|:--------|--:|-----------:|-----------:|-----------:|-----------:|
| setosa | 50 | 5.01 | 3.43 | 1.46 | 0.25 |
| versicolor | 50 | 5.94 | 2.77 | 4.26 | 1.33 |
| virginica | 50 | 6.59 | 2.97 | 5.55 | 2.03 |

The three species form visually distinct clusters in the first two PCA dimensions -- a good starting point for method comparison. For a more challenging case, add noise or reduce sample size.

---

## 5. Step 1: Fit PCA and select the number of components

```{r}
library(pcaLab)
fit <- pca_fit(iris[, 1:4], center = TRUE, scale = TRUE, ncomp = 4)
summary(fit)
fit$variance_explained[1:2]
```

The first two components capture over 95% of the total standardized variance. The `dims` argument in every clustering function lets us use any subset.

### 5.1 How many PCs to retain?

```{r}
sel <- pca_ncomp(fit, methods = c("parallel", "cv"), nperm = 49, seed = 1)
sel$recommendations
```

Parallel analysis and cross-validation both suggest 2 components -- consistent with the clear two-dimensional scatter in the score plot.

---

## 6. Step 2: Evaluate the number of clusters

`pca_cluster_evaluate()` computes WSS, silhouette, Calinski-Harabasz, and the gap statistic for each candidate k.

```{r}
ev <- pca_cluster_evaluate(iris[, 1:4], method = "kmeans", k_range = 2:6, B = 499)
ev
```

**Interpretation caution:** The gap statistic and silhouette sometimes disagree. Disagreement is informative, not a failure. Report both values and explain why you chose a particular k.

---

## 7. Step 3: Cluster the data

`pca_cluster()` accepts any supported method and returns a structured object that all downstream functions can consume.

```{r}
cl <- pca_cluster(iris[, 1:4], method = "kmeans", k = 3, dims = 1:2)
cl
```

Available methods: kmeans (base R, fast spherical), pam (cluster, robust), hclust (base R, hierarchical), mclust (model-based), dbscan (density-based).

Restrict to a subspace of principal components:

```{r}
cl2 <- pca_cluster(iris[, 1:4], method = "kmeans", k = 3, dims = 1:3)
table(cl2$cluster, iris$Species)
```

---

## 8. Step 4: Compare clustering methods

`pca_cluster_compare()` runs each method and returns a table of internal validation indices plus the adjusted Rand index between every pair.

```{r}
cmp <- pca_cluster_compare(iris[, 1:4], methods = c("kmeans", "pam", "hclust"), k = 3)
cmp
```

The Rand index measures how much two clusterings agree beyond chance. A value close to 1 means near-perfect agreement; near 0 means random overlap.

---

## 9. Step 5: Assess bootstrap stability

`pca_cluster_stability()` draws nboot bootstrap samples, clusters each sample, and constructs a consensus matrix.

```{r}
st <- pca_cluster_stability(iris[, 1:4], method = "kmeans", k = 3, nboot = 199, seed = 1)
st
```

| Diagnostic | Good range | Meaning |
|---|---|---|
| mean stability | > 0.85 | Most observations are consistently assigned |
| cophenetic correlation | > 0.75 | Consensus distances resemble original distances |
| mean silhouette | > 0.50 | Average separation between clusters |

If mean stability is below 0.70, the cluster boundaries are fragile. This is not a reason to hide the fact -- it is a scientific finding about the data.

---

## 10. Step 6: Compute cluster boundaries

`pca_cluster_boundary()` computes one of several options.

**Mahalanobis confidence ellipse (default):**

```{r}
bd <- pca_cluster_boundary(iris[, 1:4], cl$cluster, method = "ellipse", level = 0.95)
nrow(bd$boundaries)
```

**Convex hull:**

```{r}
bd_hull <- pca_cluster_boundary(iris[, 1:4], cl$cluster, method = "convex")
nrow(bd_hull$boundaries)
```

**SVM decision surface (requires e1071):**

```{r}
if (requireNamespace("e1071", quietly = TRUE)) {
  bd_svm <- pca_cluster_boundary(iris[, 1:4], cl$cluster, method = "svm")
  nrow(bd_svm$boundaries)
}
```

---

## 11. Step 7: Generate observation labels

`pca_cluster_labels()` provides four modes: group, mean, observation, and external.

```{r}
head(pca_cluster_labels(iris[, 1:4], cl$cluster, type = "group"), 10)
head(pca_cluster_labels(iris[, 1:4], cl$cluster, type = "mean"), 3)
head(pca_cluster_labels(iris[, 1:4], cl$cluster, type = "observation"), 5)
ext_lab <- c(rep("Setosa", 50), rep("Vers", 50), rep("Virgin", 50))
head(pca_cluster_labels(iris[, 1:4], cl$cluster, type = "external", labels = ext_lab), 5)
```

---

## 12. Step 8: Generate the full report

`pca_cluster_report()` gathers evaluation, stability, silhouette, and cluster sizes into a single printable summary.

```{r}
rpt <- pca_cluster_report(iris[, 1:4], method = "kmeans", k_range = 2:3)
print(rpt)
```

---

## 13. Interpretation safeguards

1. **PCA orientation is arbitrary.** Cluster boundaries depend on which PCs you keep.
2. **Stability is not validation.** High bootstrap stability does not prove that groups exist in the population.
3. **Silhouette and CH are relative indices.** They rank candidate k values, not absolute quality.
4. **Boundaries are visual aids.** Report the method and level, not just the figure.
5. **Labels are descriptive.** A cluster name based on external biology is a hypothesis, not a conclusion.

---

## 14. Putting it all together

```{r}
fit <- pca_fit(iris[, 1:4], center = TRUE, scale = TRUE)
ev <- pca_cluster_evaluate(fit, k_range = 2:5, B = 499)
cl <- pca_cluster(fit, method = "kmeans", k = ev$best_k, dims = 1:2)
st <- pca_cluster_stability(fit, method = "kmeans", k = ev$best_k, nboot = 199, seed = 1)
bd <- pca_cluster_boundary(fit, cl$cluster, method = "ellipse", level = 0.95)
lb <- pca_cluster_labels(fit, cl$cluster, type = "group")
rpt <- pca_cluster_report(fit, clustering = cl$cluster, k_range = 2:3)
cat(sprintf("k = %d | mean silhouette = %.3f | stability = %.3f | PVE = %.1f%%\n",
            cl$k, mean(cl$silhouette_per_obs, na.rm = TRUE),
            st$mean_stability, sum(fit$variance_explained[1:2]) * 100))
```

---

## 15. Notes

- All clustering functions accept either a pca_fit object or a raw numeric matrix.
- Optional engines (cluster, mclust, dbscan, e1071) are guarded with requireNamespace().
- The dims argument lets you restrict clustering to a chosen subspace.
- Bootstrap cluster stability is not a substitute for replicate study design.
