---
title: "High-Dimensional and Regularized PCA"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{High-Dimensional and Regularized PCA}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 5)
library(pcaLab)
```

## The p greater than n setting

For centered data, there are at most $n-1$ nonzero classical principal components. Sample covariance eigenvalues become unstable when $p$ is large relative to $n$.

```{r}
s <- pca_simulate(n=40,p=120,rank=4,noise=.4,seed=7)
f <- pca_fit(s$data,method="classical",scale=TRUE,ncomp=10)
f$ncomp
```

## Shrinkage PCA

With `corpcor`, `method="shrinkage"` uses a shrinkage covariance estimator before eigendecomposition.

```{r eval=FALSE}
sh <- pca_fit(s$data,method="shrinkage",scale=TRUE,ncomp=5)
```

## Randomized SVD

`method="randomized"` uses `irlba` when available and is intended for large rectangular matrices where only a small number of components is required.

## Random-matrix screening

For unit-variance independent noise, the Marchenko-Pastur upper edge is

$$\lambda_+=(1+\sqrt q)^2, \qquad q=p/(n-1).$$

`pca_ncomp(..., methods="rmt")` counts sample eigenvalues above this edge. It is a screening reference, not a universal law for dependent, heteroscedastic, or strongly non-Gaussian data.
