---
title: "Probabilistic and Bayesian PCA"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Probabilistic and Bayesian PCA}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 5)
library(pcaLab)
```

## Probabilistic PCA

PPCA models

$$x=Wz+\mu+\epsilon,$$

with $z\sim N(0,I)$ and isotropic $\epsilon\sim N(0,\sigma^2I)$. The marginal covariance is

$$C=WW^T+\sigma^2I.$$

```{r}
f <- pca_fit(iris[,1:4],method="ppca",scale=TRUE,ncomp=2)
f$extra$sigma2
f$extra$loglik
```

The maximum-likelihood loading subspace coincides with the leading classical PCA subspace under the PPCA model, while the probabilistic formulation adds a noise model and latent posterior coordinates.

## Bayesian PCA

`method="bayesian"` uses the optional `pcaMethods` BPCA backend. Bayesian credible statements depend on that model and its priors; they should not be confused with nonparametric bootstrap intervals.
