---
title: "Kernel and Nonlinear PCA"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Kernel and Nonlinear PCA}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 5)
library(pcaLab)
```

## Why linear PCA can miss curved structure

Zero or weak Pearson correlation does not imply absence of dependence. Curved manifolds can have important structure that no single linear projection summarizes well.

```{r}
s <- pca_simulate(n=150,p=5,rank=2,nonlinear=TRUE,noise=.15,seed=8)
lin <- pca_fit(s$data,scale=TRUE,ncomp=2)
k <- pca_fit(s$data,method="kernel",scale=TRUE,ncomp=2,kernel="rbf")
```

## Kernel PCA

Kernel PCA centers a Gram matrix in feature space and solves an eigenvalue problem there. `pcaLab` supports linear, polynomial, and RBF kernels and retains the training kernel quantities needed for out-of-sample projection.

```{r}
predict(k,s$data[1:5,,drop=FALSE])
```

Ordinary variable loadings do not generally exist in the original measurement space, so correlation-circle interpretation is not automatically transferred from linear PCA.

## Inverse nonlinear PCA

When `pcaMethods` is installed, `method="nlpca"` delegates to its inverse nonlinear PCA engine. Nonlinear scores are the primary result; any auxiliary loading-like output should not be interpreted as ordinary classical eigenvectors.
