---
title: "Compositional PCA"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Compositional PCA}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 5)
library(pcaLab)
```

## Compositional geometry

A composition contains relative information. Applying ordinary PCA directly to closed proportions can generate artifacts because the parts are constrained by a constant sum.

For positive compositions, the centered log-ratio transform is

$$\operatorname{clr}(x)_j=\log x_j-\frac1p\sum_{h=1}^p\log x_h.$$

```{r}
X <- matrix(c(.2,.3,.5,.1,.4,.5,.25,.25,.5,.3,.2,.5),ncol=3,byrow=TRUE)
colnames(X) <- c("N","P","K")
f <- pca_fit(X,method="compositional",coordinate="clr",ncomp=2)
f$extra$coordinate
```

## ilr coordinates

`coordinate="ilr"` uses an internal orthonormal Helmert basis and performs PCA in $(p-1)$ Euclidean log-ratio coordinates.

## Zeros

Zeros are not silently replaced. A positive `pseudocount` must be supplied explicitly if that approximation is scientifically defensible. Structural zeros and rounded zeros may require different methods outside the scope of a generic pseudocount.
