---
title: "Functional PCA"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Functional PCA}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 5)
library(pcaLab)
```

## Curves as observations

When each observational unit is a curve sampled on a common grid, functional PCA treats variation across the domain as the object of interest. The dense-grid engine uses quadrature weights to approximate the $L^2$ inner product.

```{r}
grid <- seq(0,1,length.out=60)
set.seed(1)
X <- t(vapply(1:40,function(i) sin(2*pi*grid)+rnorm(60,sd=.15)+rnorm(1)*cos(2*pi*grid),numeric(60)))
f <- pca_fit(X,method="functional",grid=grid,smooth=TRUE,ncomp=2)
head(f$extra$eigenfunctions)
```

The returned eigenfunctions are loading vectors divided by the square root of the quadrature weights.

## Multivariate functional PCA

For genuine multivariate functional objects, `backend="MFPCA"` delegates to the optional `MFPCA` package. Functional eigenfunctions are preserved in the backend-specific object and are not forced into an ordinary variable-loading matrix.
