---
title: "Dynamic PCA for Multivariate Time Series"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Dynamic PCA for Multivariate Time Series}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 5)
library(pcaLab)
```

## Lag embedding

Dynamic PCA augments the current multivariate vector with lagged vectors before fitting PCA. This allows components to represent contemporaneous and lagged covariance structure.

```{r}
set.seed(1)
X <- matrix(arima.sim(list(ar=.7),n=400),100,4)
f <- pca_fit(X,method="dynamic",lags=c(1,2),scale=TRUE,ncomp=3)
colnames(f$processed_data)[1:8]
```

## Time ordering matters

Ordinary observation bootstrap is generally not appropriate when serial dependence is inferentially relevant. Use time-series-aware resampling outside the generic `pca_boot()` workflow or predefine independent experimental units.

## Robust dynamic PCA

If `rospca` is available, `robust=TRUE` fits ROBPCA to the lag-embedded matrix.

```{r eval=FALSE}
r <- pca_fit(X,method="dynamic",lags=1:2,robust=TRUE,scale=TRUE,ncomp=3)
```
