---
title: "Multilevel PCA, ASCA, and Multiblock PCA"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Multilevel PCA, ASCA, and Multiblock PCA}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 5)
library(pcaLab)
```

## Repeated measurements and multilevel decomposition

For observations nested within subjects, `method="multilevel"` can fit PCA to within-subject deviations or between-subject means.

```{r}
X <- as.matrix(iris[,1:4])
subject <- rep(1:30,length.out=nrow(X))
w <- pca_fit(X,method="multilevel",subject=subject,level="within",scale=TRUE,ncomp=2)
b <- pca_fit(X,method="multilevel",subject=subject,level="between",scale=TRUE,ncomp=2)
```

## ASCA for designed experiments

ASCA first decomposes the multivariate response according to a linear-model design and then applies PCA to effect matrices.

```{r}
set.seed(1)
d <- data.frame(A=factor(rep(1:2,each=12)),B=factor(rep(1:3,8)))
Y <- cbind(y1=rnorm(24)+2*as.numeric(d$A), y2=rnorm(24)+as.numeric(d$B), y3=rnorm(24))
a <- pca_asca(Y,~A+B,d,ncomp=2)
a$effect_ss
```

## Multiblock PCA

Blocks can be scaled by Frobenius norm or first singular value before concatenation, preventing a large block from dominating only because it contains more variables or greater total energy.

```{r}
mb <- pca_multiblock(list(morph=iris[,1:2],size=iris[,3:4]),ncomp=2,scale=TRUE)
mb$extra$block_contributions
```

## Streaming (incremental) PCA

For data that arrives in chunks, the incremental API accumulates running moments and finalizes an eigendecomposition without storing every observation:

```{r}
set.seed(9)
X <- matrix(rnorm(600), 150, 4)
st <- pca_incremental_init(4, center = TRUE, scale = TRUE)
st <- pca_incremental_update(st, X[1:50, ])
st <- pca_incremental_update(st, X[51:100, ])
st <- pca_incremental_update(st, X[101:150, ])
inc <- pca_incremental_finalize(st, ncomp = 2)
inc$eigenvalues
# compare with the batch eigendecomposition
ref <- eigen(cov(scale(X, center = TRUE, scale = TRUE)), symmetric = TRUE)$values
max(abs(inc$eigenvalues - ref[1:2]))
```
