---
title: "Projection Tours Anchored to PCA"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Projection Tours Anchored to PCA}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 5)
library(pcaLab)
```

## Why tour after PCA?

A two-dimensional PCA plot is only one projection. A local tour explores nearby projections to assess visual stability; a grand tour explores a broad set of projections; a guided tour optimizes a projection pursuit index.

```{r eval=FALSE}
fit <- pca_fit(iris[,1:4],scale=TRUE,ncomp=4)
pca_tour(fit,type="local",group=iris$Species)
pca_tour(fit,type="guided",group=iris$Species)
h <- pca_tour(fit,type="history",save_history=TRUE)
```

## Interpret tours as exploration

A projection discovered by a guided tour can reveal clustering or non-Gaussian structure that variance-maximizing PCA does not emphasize. This does not invalidate PCA; it means the projection objective is different.

## Compare tour histories

```{r}
fit <- pca_fit(iris[,1:4],scale=TRUE,ncomp=4)
tc <- pca_tour_compare(fit, max_bases = 10)
length(tc$local)
length(tc$grand)
```

`pca_tour_compare()` stores PCA, local, guided, and grand-tour histories for side-by-side investigation. The local history quantifies how much the PCA view changes under nearby rotations; the grand history samples the full projection space.
