---
title: "When PCA Is Not the Right Tool"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{When PCA Is Not the Right Tool}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 5)
library(pcaLab)
```

## Weak correlation or isotropic structure

If variables are nearly uncorrelated and have similar standardized variance, no small set of linear directions may dominate. A flat eigenspectrum is an informative result rather than a software failure.

## Curved manifolds

Linear PCA may miss nonlinear dependence. Compare kernel or nonlinear methods only when the scientific objective supports nonlinear coordinates.

## Discrete outcomes

Binary/count data can require logistic or generalized PCA rather than Gaussian covariance PCA.

## Compositions

Closed proportions should usually be analyzed in log-ratio geometry rather than directly.

## Repeated, temporal, functional, and multiblock data

Ignoring data structure can produce misleading uncertainty and component interpretation. Use the corresponding structured methods when the design requires them.

## Supervised separation is a different objective

PCA is not a classifier and does not seek group separation. Coloring score plots by treatment after fitting is descriptive. If the scientific question is a treatment effect, use a design-based multivariate inferential model, ASCA/permutation design, discriminant method, mixed model, or other approach appropriate to the experiment.

```{r}
ex <- pca_example_agronomy()
d <- pca_doctor(ex$data)
d$warnings
d$suggestions
```

## A practical decision rule

Use PCA when the scientific question concerns dominant variance/covariance structure and a linear low-dimensional representation is meaningful. Select another method because its estimand matches the scientific question, not because it gives a visually cleaner plot.
